Human Factors Influencing Visual Statistical Inference

### 2013-07-18 sociology_paper.tex
reorganised the sections

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Freedom's Lighthouse Averages \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   RealClearPolitics Averages\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms of 10,000 simulations each based on polling averages from two different sources. For the histogram on the top, the $p$-value of observing results as extreme as  the  2012 U.S. election results based on the bootstrap is 0.0533 (with Bootstrap standard error of 0.002), making the election results almost significantly different from the polls. There is no indication of any inconsistency between polls and election results based on the bootstrap simulation below. The lineups are based on the top source. }
\end{figure}

@@

\begin{table*}[hbtp] 
\centering 
\caption{Visual test statistics used in 10 different simulation experiments. The observers are asked different questions to answer while evaluating a lineup. All the experiment data are used to estimate the effect of demographic factors on visual inference while some are used to estimate the learning trend and location effect of actual data in the lineup.} 
\begin{tabular}{m{.5cm}m{2.6cm}m{2cm}m{4.5cm}m{3.5cm}} 
\hline\hline 
Serial & Experiment &  Test Statistic  & Lineup question & Provided data to estimate\\ [0.5ex] % inserts table %heading 
\hline 
1  & Box plot & \begin{minipage}[t]{2cm} \begin{center}	\scalebox{0.12}{\includegraphics{stat_category1.pdf}} \end{center} \end{minipage} & Which set of box plots shows biggest vertical difference 
between group A and B? & Demographic factor\\
2 &  Scatter plot & \begin{minipage}[t]{2cm}  \begin{center} \scalebox{0.3}{\includegraphics{stat_beta_k1.png}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? & Demographic factor\\
  3 & Contaminated plot &\begin{minipage}[t]{2cm} \begin{center} \scalebox{0.5}{\includegraphics{stat_contaminated.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? & Demographic factor\\
 4 & Polar vs Cartesian & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.32}{\includegraphics{stat_polar.pdf}} \end{center} \end{minipage} &Which plot is different?& Demographic factor\\
  5 & Hist vs density & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.38}{\includegraphics{stat_density.pdf}} \end{center} \end{minipage} &In which plot is the blue group furthest to the right?& Learning trend and Demographic factor\\  
  6 & Violin vs boxplot & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_violin.pdf}} \end{center} \end{minipage}&  In which plot does the blue group look the most different from  the red group? & Learning trend and Demographic factor\\
  7 & Group separation & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.4}{\includegraphics{stat_separation.pdf}} \end{center} \end{minipage} &Which of these plots has the most separation between the coloured groups? & Learning trend and  Demographic factor\\ 
  8 & Sine Illusion & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.28}{\includegraphics{stat_sine.pdf}} \end{center} \end{minipage} & In what picture is the size of the curve most consistent? & Demographic factor\\ 
  9 & Gene expression &\begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.45}{\includegraphics{stat_gene.pdf}} \end{center} \end{minipage} & In which of these plots is the green line the steepest, and the spread of the green points relatively small? & Location effect and Demographic factor\\ 
  10 & Test normality & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_normality.pdf}} \end{center} \end{minipage} &Which of these plots is most different from the others? & Demographic factor\\

@@

\section{Factors Affecting Visual Inference} \label{sec:factor_performance}

@@

\section{Experimental Designs and Methods}\label{sec:exp_design}
While evaluating a lineup, the performance of the observer may depend on the factors described in Section \ref{sec:factor_performance}. Some of these factors such as signal in the data and individual abilities were studied in \cite{majumder:2013}. In this paper we intend to study effect of the factors such as human demographics, learning trend, actual plot locations and selection of null plots. This section presents the simulation experiment setup, data collection methods and data analysis plans with models to asses the influence of those factors on visual statistical inference.

@@

\subsection{Experiment Setup}  In addition to obtain evaluation responses of the lineups, all the 10 experiments shown in Table \ref{tbl:visual_stat} were designed to collect the following demographic information of the subjects participating the experiments:

@@

%Multiple simulation experiments are designed to investigate the learning trend of the observer and the location effect of the actual data in the lineup. These are explained in tis section.

\subsubsection{Learning Trend} Learning trend of a subject can be observed in terms of performance over successive feedbacks received when multiple lineups are shown for evaluation. Experiments 5, 6 and 7 were used for this. Each subject was shown a total of 10 lineups randomly selected from a pool of lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups were randomized. The responses of the lineups were recorded by attempt 1 through 10. Attempt 1 means that the response is for the first lineup the observer evaluates and attempt 10 refers to the response for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\subsubsection{Design for Location Effect} \label{sec:location_design} Experiment 9 shown in Table~\ref{tbl:visual_stat} is designed to to assess the location effect of a data plot in a lineup. It is set up based on the gene expression data \citep{atwood:2013} with two groups. One is the Genotype and the other the Empty Vector (EV). For each of the groups, gene expression data were collected in presence or absence of iron sufficiency. Two factors were of primary interest. One is the Genotype main effect and the other is the interaction effect between Genotype and iron sufficiency.

@@

%We set up a Turk experiment to examine whether there is any difference of performance based on the location of actual data plot in the lineup. For this five locations of a lineup were randomly chosen to put the same actual data plot. The locations are 2,9,12,16,20 for Interaction effect and 1,8,12,17,20 for Genotype effect. For a lineup with specific data plot location, five sets of null plots are used giving 5 lineups for each location. In total we have 25 lineups for Interaction effect and 25 lineups for Genotype effect. These 50 lineups are evaluated by 100 people recruited from Amazon Mechanical Turk. Each person evaluated 3 lineups one from Interaction, one from Genotype and the third one was a test lineup. The test lineup is used to process the payment and examine the quality of the data and the responses of the test plot is not added to our analysis.

%One such lineup is shown in Figure \ref{fig:lineup_geno}.

@@

@@

\subsection{Model to Estimate Demographic Factor Effect} The feedback provided by each observer on a lineup is a binary response variable. Suppose $Y_{ij}$ denotes the response from observer $i$ on a  lineup $j$. $Y_{ij} =1$ if the response is correct otherwise $Y_{ij} =0$. Let $\pi_{ij}=  E(Y_{ij})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$. We model this in a generalized mixed effects model of the form
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\mu$ is an overall average probability for picking out the data plot from a lineup. $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender category $l(i)$, education level $m(i)$ and country name $s(i)$ respectively for observer $i$. $\ell_j$ is a random intercept predicting lineup difficulty level and we assume  $\ell_j \sim N(0, \sigma_\ell^2)$. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

% and $k(i) \in \{18-25, 26-30, 31-35, 36-40, 41-45, 46-50, \text{above } 50 \}$. $\gamma$ is the effect of gender category $l(i) \in \{\text{Male, Female}\}$. $\tau$ is the effect of education levels $m(i) \in \{ \text{High school, Undergrad course, Undergrad degree, Graduate course, }\}$

\subsection{Model to Estimate Learning Trend}

Multiple lineups were sequentially shown to each observer for evaluation. Suppose $Y_{ijk}$ denotes the response from observer $i$ on a  lineup $j$ in his/her $k$th attempt.  $Y_{ijk}=1$ if the response is correct otherwise  $Y_{ijk}=0$. Let $\pi_{ijk}=  E(Y_{ijk})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$ in his/her $k$th attempt. We model this in a generalized mixed effects model of the form
\begin{equation} \label{eqn:trend_response}
g( \pi_{ijk} )= \mu + \alpha_k + u_i +  a_{i} k + \ell_j,  
\end{equation}
where $\mu$ is an overall average probability for picking out the data plot from a lineup, $\alpha_k$ is the effect of the $k$th attempt on the probability, with $\alpha_1 = 0$ and $k = 1, ..., K$.
$u_i$ and $a_i$ are observer specific random effects, $i = 1, ..., I$. $u_i$ is a random intercept, describing a basic subject-specific ability. We assume $u_i \sim N(0, \sigma_u^2)$. 
$a_i$ is a random slope capturing an individual's specific learning effect over the course of $K$ attempts. We assume $a_i \sim N(0, \sigma_a^2)$. 
For $\ell_j$ we again assume a normal distribution, $N(0, \sigma_\ell^2)$. $\ell_j$ is a random intercept predicting lineup difficulty level. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$ .

%$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well as the attempt made by the observer. $Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
%$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups.

The inverse link function, $g^{-1}(.)$, from  Equation \ref{eqn:trend_response} leads to the estimate of the subject and the lineup specific probability of successful evaluation in $kth$ attempt by a single observer as 
\begin{equation} \label{eqn:trend_power}
\hat p_{ijk} =  g^{-1}(\hat{\mu} + \hat{\alpha}_k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j).
\end{equation}

The learning of each observer over a course of $K$ evaluations may be observed as the improvement of the time taken to evaluate a lineup in the later attempts. Suppose $Z_{ijk}$ denotes the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$ in his/her $k$th attempt. Let $\mu_{ijk}=  E(Z_{ijk})$ be the average of log(time taken) by  observer $i$ to pick the data panel from lineup $j$ in his/her $k$th attempt. We model this in a mixed effects model of the form
%\hh{
\begin{equation} \label{eqn:trend_time}
Z_{ijk} = \mu + \alpha_1 + \alpha k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup. $\alpha$ is the average change in log time taken for each additional attempt.  $\alpha_1$ is an offset in log time taken for the first attempt. All other effects are random effects: as before, $u_i$ is a subject-specific intercept representing  individual speed of an observer with $u_i \sim N(0, \sigma_u^2)$. $a_i$ is a subject-specific slope representing the deviation of the speed-up (or -down) by attempt $k$. We assume $a_i \sim N(0, \sigma_a^2)$. $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup; $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.
%}

%\begin{equation} \label{eqn:trend_time}
%Z_{ijk} = \mu_1 + \alpha_k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
%\end{equation}
%where the link function $g_1(\mu)=1/\mu;  \mu >0$ and $\mu_1$ represents the inverse of overall average time taken to evaluate a lineup.

%For $Y_{l,i}$ being binary response we use {\it logit} link function $g(\mu)=g_1(\mu)=\log(\mu) - \log(1-\mu); 0 \le \mu \le 1$ and $\hat p_{\ell i}$ gives the estimate of proportion correct by subject $i$ with lineup $\ell$.  The response time taken is a positive random variable with a right skewed distribution. For this the usual convention is to model a gamma distribution with time. Thus we used inverse link function $g(\mu)=g_2(\mu)=1/\mu;  \mu >0$ for gamma response $Y_{l,i}$ and $\hat p_{\ell i}$ gives the estimate of mean time taken by subject $i$ for lineup $\ell$.

\subsection{Model to Estimate Location Effect}

>>>>

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms of 10,000 simulations each based on polling averages from two different sources. For the histogram on the top, the $p$-value of observing results as extreme as  the  2012 U.S. election results based on the bootstrap is 0.0533 (with Bootstrap standard error of 0.002), making the election results almost significantly different from the polls (data source for poll aggregator 1 is Freedom`s Lighthouse Averages). There is no indication of any inconsistency between polls and election results based on the bootstrap simulation below (data source for poll aggregator 2 is RealClearPolitics Averages). The lineups are based on the top source.}
\end{figure}

@@

\begin{table*}[hbtp] 
\centering 
\caption{Visual test statistics used in 10 different simulation experiments. The observers are asked different questions to answer while evaluating a lineup. All the experiment data are used to estimate the effect of demographic factors (DF) on visual inference while some are used to estimate the learning trend (LT) and location effect (LE) of actual data in the lineup as shown in the right most column.} 
\begin{tabular}{m{.5cm}m{2.6cm}m{2cm}m{5.5cm}m{3cm}} 
\hline\hline 
Serial & Experiment &  Test Statistic  & Lineup question & Designed to study\\ [0.5ex] % inserts table %heading 
\hline 
1  & Box plot & \begin{minipage}[t]{2cm} \begin{center}	\scalebox{0.12}{\includegraphics{stat_category1.pdf}} \end{center} \end{minipage} & Which set of box plots shows biggest vertical difference 
between group A and B? & DF\\
2 &  Scatter plot & \begin{minipage}[t]{2cm}  \begin{center} \scalebox{0.3}{\includegraphics{stat_beta_k1.png}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? & DF\\
  3 & Contaminated plot &\begin{minipage}[t]{2cm} \begin{center} \scalebox{0.5}{\includegraphics{stat_contaminated.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? & DF\\
 4 & Polar vs Cartesian & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.32}{\includegraphics{stat_polar.pdf}} \end{center} \end{minipage} &Which plot is different?& DF\\
  5 & Hist vs density & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.38}{\includegraphics{stat_density.pdf}} \end{center} \end{minipage} &In which plot is the blue group furthest to the right?&  DF and LT \\  
  6 & Violin vs boxplot & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_violin.pdf}} \end{center} \end{minipage}&  In which plot does the blue group look the most different from  the red group? &  DF and LT \\
  7 & Group separation & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.4}{\includegraphics{stat_separation.pdf}} \end{center} \end{minipage} &Which of these plots has the most separation between the coloured groups? & DF and LT \\ 
  8 & Sine Illusion & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.28}{\includegraphics{stat_sine.pdf}} \end{center} \end{minipage} & In what picture is the size of the curve most consistent? & DF\\ 
  9 & Gene expression &\begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.45}{\includegraphics{stat_gene.pdf}} \end{center} \end{minipage} & In which of these plots is the green line the steepest, and the spread of the green points relatively small? & DF and LE\\ 
  10 & Test normality & \begin{minipage}[t]{2cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_normality.pdf}} \end{center} \end{minipage} &Which of these plots is most different from the others? & DF\\

@@

\section{Factors Potentially Affecting Visual Inference} \label{sec:factor_performance}

@@

\section{Experimental Methods}\label{sec:exp_design}
While evaluating a lineup, the performance of the observer may depend on the factors described in Section \ref{sec:factor_performance}. Some of these factors such as signal in the data and individual abilities were studied in \cite{majumder:2013}. In this paper we intend to study effect of the factors such as human demographics, learning trend, actual plot locations and selection of null plots. This section presents the simulation experiment setup, data collection methods and data analysis plans with models to asses the influence of those factors on visual statistical inference.

@@

\subsection{Demographic Factors}  In addition to obtain evaluation responses of the lineups, all the 10 experiments shown in Table \ref{tbl:visual_stat} were designed to collect the following demographic information of the subjects participating the experiments:

@@

The feedback provided by each observer on a lineup is a binary response variable. Suppose $Y_{ij}$ denotes the response from observer $i$ on a  lineup $j$. $Y_{ij} =1$ if the response is correct otherwise $Y_{ij} =0$. Let $\pi_{ij}=  E(Y_{ij})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$. We model this in a generalized mixed effects model of the form
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\mu$ is an overall average probability for picking out the data plot from a lineup. $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender category $l(i)$, education level $m(i)$ and country name $s(i)$ respectively for observer $i$. $\ell_j$ is a random intercept predicting lineup difficulty level and we assume  $\ell_j \sim N(0, \sigma_\ell^2)$. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

\subsection{Learning Trend} Learning trend of a subject can be observed in terms of performance over successive feedbacks received when multiple lineups are shown for evaluation. Experiments 5, 6 and 7 were used for this. Each subject was shown a total of 10 lineups randomly selected from a pool of lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups were randomized. The responses of the lineups were recorded by attempt 1 through 10. Attempt 1 means that the response is for the first lineup the observer evaluates and attempt 10 refers to the response for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

%\subsection{Model to Estimate Learning Trend}

Multiple lineups were sequentially shown to each observer for evaluation. Suppose $Y_{ijk}$ denotes the response from observer $i$ on a  lineup $j$ in his/her $k^{th}$ attempt.  $Y_{ijk}=1$ if the response is correct otherwise  $Y_{ijk}=0$. Let $\pi_{ijk}=  E(Y_{ijk})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$ in his/her $k^{th}$ attempt. We model this in a generalized mixed effects model of the form
\begin{equation} \label{eqn:trend_response}
g( \pi_{ijk} )= \mu + \alpha_k + u_i +  a_{i} k + \ell_j,  
\end{equation}
where $\mu$ is an overall average probability for picking out the data plot from a lineup, $\alpha_k$ is the effect of the $k^{th}$ attempt on the probability, with $\alpha_1 = 0$ and $k = 1, ..., K$.
$u_i$ and $a_i$ are observer specific random effects, $i = 1, ..., I$. $u_i$ is a random intercept, describing a basic subject-specific ability. We assume $u_i \sim N(0, \sigma_u^2)$. 
$a_i$ is a random slope capturing an individual's specific learning effect over the course of $K$ attempts. We assume $a_i \sim N(0, \sigma_a^2)$. 
For $\ell_j$ we again assume a normal distribution, $N(0, \sigma_\ell^2)$. $\ell_j$ is a random intercept predicting lineup difficulty level. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$ .

The inverse link function, $g^{-1}(.)$, from  Equation \ref{eqn:trend_response} leads to the estimate of the subject and the lineup specific probability of successful evaluation in $k^{th}$ attempt by a single observer as 
\begin{equation} \label{eqn:trend_power}
\hat p_{ijk} =  g^{-1}(\hat{\mu} + \hat{\alpha}_k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j).
\end{equation}

The learning of each observer over a course of $K$ evaluations may be observed as the improvement of the time taken to evaluate a lineup in the later attempts. Suppose $Z_{ijk}$ denotes the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$ in his/her $k^{th}$ attempt. Let $\mu_{ijk}=  E(Z_{ijk})$ be the average of log(time taken) by  observer $i$ to pick the data panel from lineup $j$ in his/her $k^{th}$ attempt. We model this in a mixed effects model of the form
%\hh{
\begin{equation} \label{eqn:trend_time}
Z_{ijk} = \mu + \alpha_1 + \alpha k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup. $\alpha$ is the average change in log time taken for each additional attempt.  $\alpha_1$ is an offset in log time taken for the first attempt. All other effects are random effects: as before, $u_i$ is a subject-specific intercept representing  individual speed of an observer with $u_i \sim N(0, \sigma_u^2)$. $a_i$ is a subject-specific slope representing the deviation of the speed-up (or -down) by attempt $k$. We assume $a_i \sim N(0, \sigma_a^2)$. $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup; $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.

\subsection{Location Effect} \label{sec:location_design} Experiment 9 shown in Table~\ref{tbl:visual_stat} is designed to to assess the location effect of a data plot in a lineup. It is set up based on the gene expression data \citep{atwood:2013} with two groups. One is the Genotype and the other the Empty Vector (EV). For each of the groups, gene expression data were collected in presence or absence of iron sufficiency. Two factors were of primary interest. One is the Genotype main effect and the other is the interaction effect between Genotype and iron sufficiency.

@@

@@

%\subsection{Model to Estimate Location Effect}

@@

%\subsection{Model to Estimate Demographic Factor Effect} The feedback provided by each observer on a lineup is a binary response variable. Suppose $Y_{ij}$ denotes the response from observer $i$ on a  lineup $j$. $Y_{ij} =1$ if the response is correct otherwise $Y_{ij} =0$. Let $\pi_{ij}=  E(Y_{ij})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$. We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:demographic_response}
%g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
%\end{equation}
%where $\mu$ is an overall average probability for picking out the data plot from a lineup. $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender category $l(i)$, education level $m(i)$ and country name $s(i)$ respectively for observer $i$. $\ell_j$ is a random intercept predicting lineup difficulty level and we assume  $\ell_j \sim N(0, \sigma_\ell^2)$. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
%
%
%The effect of demographic factors can be observed as time taken for each evaluation by an observer. Suppose $Z_{ij}$ denotes the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$. Let $\mu_{ij}=  E(Z_{ij})$ be the average of log(time taken) by  observer $i$ to pick the data panel from lineup $j$. We model this in a mixed effects model of the form
%%\hh{
%\begin{equation} \label{eqn:demographic_time}
%Z_{ij} = \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j+ \epsilon_{ij},  
%\end{equation}
%where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup. $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are as described in Model \eqref{eqn:demographic_response}. $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup; $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.

%\subsection{Model to Estimate Learning Trend}
%
%Multiple lineups were sequentially shown to each observer for evaluation. Suppose $Y_{ijk}$ denotes the response from observer $i$ on a  lineup $j$ in his/her $k^{th}$ attempt.  $Y_{ijk}=1$ if the response is correct otherwise  $Y_{ijk}=0$. Let $\pi_{ijk}=  E(Y_{ijk})$ be the probability that  observer $i$ correctly picks the data panel from lineup $j$ in his/her $k^{th}$ attempt. We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:trend_response}
%g( \pi_{ijk} )= \mu + \alpha_k + u_i +  a_{i} k + \ell_j,  
%\end{equation}
%where $\mu$ is an overall average probability for picking out the data plot from a lineup, $\alpha_k$ is the effect of the $k^{th}$ attempt on the probability, with $\alpha_1 = 0$ and $k = 1, ..., K$.
%$u_i$ and $a_i$ are observer specific random effects, $i = 1, ..., I$. $u_i$ is a random intercept, describing a basic subject-specific ability. We assume $u_i \sim N(0, \sigma_u^2)$. 
%$a_i$ is a random slope capturing an individual's specific learning effect over the course of $K$ attempts. We assume $a_i \sim N(0, \sigma_a^2)$. 
%For $\ell_j$ we again assume a normal distribution, $N(0, \sigma_\ell^2)$. $\ell_j$ is a random intercept predicting lineup difficulty level. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$ . 
%
%The inverse link function, $g^{-1}(.)$, from  Equation \ref{eqn:trend_response} leads to the estimate of the subject and the lineup specific probability of successful evaluation in $k^{th}$ attempt by a single observer as 
%\begin{equation} \label{eqn:trend_power}
%\hat p_{ijk} =  g^{-1}(\hat{\mu} + \hat{\alpha}_k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j).
%\end{equation}
%
%The learning of each observer over a course of $K$ evaluations may be observed as the improvement of the time taken to evaluate a lineup in the later attempts. Suppose $Z_{ijk}$ denotes the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$ in his/her $k^{th}$ attempt. Let $\mu_{ijk}=  E(Z_{ijk})$ be the average of log(time taken) by  observer $i$ to pick the data panel from lineup $j$ in his/her $k^{th}$ attempt. We model this in a mixed effects model of the form
%%\hh{
%\begin{equation} \label{eqn:trend_time}
%Z_{ijk} = \mu + \alpha_1 + \alpha k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
%\end{equation}
%where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup. $\alpha$ is the average change in log time taken for each additional attempt.  $\alpha_1$ is an offset in log time taken for the first attempt. All other effects are random effects: as before, $u_i$ is a subject-specific intercept representing  individual speed of an observer with $u_i \sim N(0, \sigma_u^2)$. $a_i$ is a subject-specific slope representing the deviation of the speed-up (or -down) by attempt $k$. We assume $a_i \sim N(0, \sigma_a^2)$. $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup; $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.
%
%
%Equation \ref{eqn:trend_time} leads to the estimate of the subject and the lineup specific time taken for an evaluation in $kth$ attempt by a single observer as 
%\begin{equation} \label{eqn:trend_time_est}
%\hat \mu_{ijk} =  \hat{\mu} + \hat{\alpha_1}+ \hat{\alpha}k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j.
%\end{equation}
%
%To fit all these mixed effect models the function $lmer()$ is used from R package $lme4$ by \cite{lme4:2011}. To obtain the $p$-values of fixed effect parameters estimates, normal approximation is used for $Z$ scores computed as the ratio of estimates to the estimated standard errors.

%\subsection{Model to Estimate Location Effect}
%
%
%As per the design of experiment for location effect, the actual data plot is same for each of the null sets of plot. Thus the response data of this experiment constitute a multivariate response. To examine if the difference in proportion correct among the locations is statistically significant we fit a one way multi variate analysis of variance (MANOVA) model to the data.
%
%Suppose $\mathbf{Y}=(Y_1,Y_2, ... , Y_p)^\top$ be a vector of random variable with dimension $p$, the total number of null sets. let $\mathbf{Y}_{ij}$ represents $jth$ vector response for $ith$ location with $i=1,2, ..., I$. We fit the following MANOVA model 
%\begin{equation}\label{manova}
%\mathbf{Y}_{ij} = \mu_{i} + \epsilon_{ij}
%\end{equation}
%where $\mu_{i}= (\mu_{1i},\mu_{2i}, ..., \mu_{pi})^\top$ is the mean vector for location $i$ and $Var(\epsilon_{ij})=\Sigma$.
